Mining the human microbiome for clinical insight

### 2018-12-12 donor-selection/ms.tex
Mostly cosmetic changes, mostly singlespace bibs

Given the availability of microbiome data from completed FMT studies, we tested this hypothesis that IBD trials would benefit from a rational donor selection strategy based on the ``absence of beneficial taxa'' disease model.
We re-analyzed microbiome data from three completed IBD FMT trials which provided publicly available sequencing data for patient and donor samples \cite{Kump2018,Goyal2018,Jacob2017}.
We selected butyrate-producers based on their genus-level taxonomy, using the genera identified in Vital et al. 2017 (see Methods, \cite{Vital2017}).
Donors in the three studies exhibited a range of total abundances of butyrate-producing bacteria (Figure \ref{fig:ibd-butyrate}A).
Surprisingly, however, the abundance of butyrate producers in the donor stool was not associated with recipient patients' clinical responses (Figure \ref{fig:ibd-butyrate}B).
We also found no association with response when matching donor abundances with their respective patient's original abundance of butyrate producers (Supplementary Figure \ref{fig:ibd-butyrate-patient-matching}).
These results show that selecting donors based on the abundance of butyrate producers may not yield improved clinical trial outcomes in IBD, and illustrates the process by which clinicians could approach and validate a rational donor selection strategy based on individual taxa.
More complex methods to identify butyrate producers (e.g. using phylogenetic-aware methods and/or metagenomics data) could be used in the next iteration to develop a donor selection strategy, if these data are available to clinicians.
Another approach, discussed below, is to select donors based on functional community assays and direct measurement of butyrate production rather than microbial taxonomies alone.

@@

Some microbiome-associated diseases may not be addressable by replenishing the patient with a generically healthy community or by targeting individual taxa, and may instead be mediated by the microbiome through a community-level function.
Here, there may not be a consistent disease-associated microbiome across patients in terms of taxonomic composition, but patients may be characterized by having microbiomes which are similarly missing or enriched in some core functionality.
This model may also apply to conditions where there are consistent disease- or health-associated taxa, but in which their collective functioning is the more important mediator of disease.
The IBD case study described above may reflect this situation: although depletion of butyrate producers is strongly associated with IBD throughout the literature, a successful donor selection strategy may need to consider butyrate production directly rather than through the proxy of taxonomy \cite{Duvallet2017,Schrimer2018}.

@@

>>>>

Given the availability of microbiome data from completed FMT studies, we tested this hypothesis that IBD trials would benefit from a rational donor selection strategy based on the ``absence of beneficial taxa'' disease model.
We re-analyzed microbiome data from three completed IBD FMT trials which provided publicly available sequencing data for patient and donor samples \cite{Kump2018,Goyal2018,Jacob2017}.
We selected butyrate-producers based on their genus-level taxonomy, using the genera identified in Vital et al. 2017 (see Methods, \cite{Vital2017}).
Donors in the three studies exhibited a range of total abundances of butyrate-producing bacteria (Figure \ref{fig:ibd-butyrate}A).
Surprisingly, however, the abundance of butyrate producers in the donor stool was not associated with recipient patients' clinical responses (Figure \ref{fig:ibd-butyrate}B).
We also found no association with response when matching donor abundances with their respective patient's original abundance of butyrate producers (Supplementary Figure \ref{fig:delta-butyrate}).
These results show that selecting donors based on the abundance of butyrate producers may not yield improved clinical trial outcomes in IBD, and illustrates the process by which clinicians could approach and validate a rational donor selection strategy based on individual taxa.
More complex methods to identify butyrate producers (e.g. using phylogenetic-aware methods and/or metagenomics data) could be used in the next iteration to develop a donor selection strategy, if these data are available to clinicians.
Another approach, discussed below, is to select donors based on functional community assays and direct measurement of butyrate production rather than microbial taxonomies alone.

@@

Some microbiome-associated diseases may not be addressable by replenishing the patient with a generically healthy community or by targeting individual taxa, and may instead be mediated by the microbiome through a community-level function.
Here, there may not be a consistent disease-associated microbiome across patients in terms of taxonomic composition, but patients may be characterized by having microbiomes which are similarly missing or enriched in some core functionality.
This model may also apply to conditions where there are consistent disease- or health-associated taxa, but in which their collective functioning is the more important mediator of disease.
The IBD case study described above may reflect this situation: although depletion of butyrate producers is strongly associated with IBD throughout the literature, a successful donor selection strategy may need to consider butyrate production directly rather than through the proxy of taxonomy \cite{Duvallet2017,Schirmer2018}.

@@

\section{Supplementary Figures}
\FloatBarrier
\begin{figure}[h]
    \begin{center}
    \includegraphics[width=\textwidth]{suppfig_delta_butyrate_vs_response.png}
    \caption{Difference between abundance of butyrate producers in donor sample and respective patient sample, stratified by patient response.}\label{fig:delta-butyrate}
    \end{center}
\end{figure}
